---
title: "Unit 11 Homework"
author: "Dana Geislinger"
date: "April 2, 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=TRUE, fig.height=6, fig.width=8)
```

### **1:** Brief Financial Data
We will be analyzing annual European Stock Data from 1990-1998. Particularly, we are interested in charting the prices for the DAX index and decomposing that multiplicative time series into separate trend, seasonality, and random plots.

```{r 1}
# Plot data for DAX in blue and draw a red to indicate 1997
plot(EuStockMarkets[, 'DAX'],
     main='Daily Closing Stock Prices for EU Index DAX from 1991 to 1998',
     xlab='Year',
     ylab='DAX Index Daily Closing Price',
     col='blue'
     )
abline(v=1997, col='red')

# Decompose DAX timeseries and plot trend, seasonality and random
dax.decomp = decompose(EuStockMarkets[, 'DAX'], type='mult')
plot(dax.decomp,
     col='blue'
     )
abline(v=1997, col='red')
```

### **2:** Temperature Data
We are interested in predicting a future trend for the maximum annual temperature at Moorabbin Airport in Melbourne based on historical data spanning 1971-2016. Multiple prediction models will be compared to find the ideal model suited for this problem.
```{r 2}
# Load necessary libraries for data and tools
suppressMessages(require("fpp2"))
suppressMessages(require("dygraphs"))
suppressMessages(require("xts"))
suppressMessages(require("forecast"))

# Plot the temperature time series
autoplot(maxtemp,
         xlab="Year",
         ylab="Temperature (Celsius)"
         )

# Subset the timeseries from 1990-2016
ts<-window(maxtemp, start=1990)

# Use SES
fit<-ses(ts, h=5)

# Compare Forecast Fit
plot(fit,
     ylab="Temperature (Celsius)",
     xlab= "Year",
     main="Comparing Forecast Fit"
     )
lines(fitted(fit), col="blue")
lines(fit$mean, col="blue", type="o")

# Find AICc
fit$model

# Fit damped Holt Fit and plot
holtfit<- holt(ts, initial='optimal', h=5, damped=TRUE)

plot(holtfit,
     ylab="Temperature (Celsius)",
     xlab= "Year",
     main="Comparing Forecast Fit"
     )
lines(fitted(holtfit), col="blue", type="o")
lines(holtfit$mean, col="red")

# Print AICc
holtfit$model
```

The SES AICc value is lower than the damped Holt's AICc value, so the SES model provides a better fit as a predictor for the temperature data.

### **3:** The Wands Choose the Wizard
We would like to interpret wand sales data from two wand salesmen over the period 1970-2017. To do this, we will import, clean, and merge the two sales data sets in R and create an interactive DyGraph of the combined data.

```{r 3}
# Import data as 2 dataframes
olliv.df = read.csv('Unit11TimeSeries_Ollivander.csv', header=FALSE)
gregor.df = read.csv('Unit11TimeSeries_Gregorovitch.csv', header=FALSE)

# Create meaningful variable names
names(olliv.df) = c("Date", "Sold_Ollivander")
names(gregor.df) = c("Date", "Sold_Gregorovitch")

# Convert 'Date' column to datetime objects
olliv.df$Date = as.Date(olliv.df$Date, format="%m/%d/%Y")
gregor.df$Date = as.Date(gregor.df$Date, format="%m/%d/%Y")

# Convert dataframes to 'xts' timeseries objects
olliv.ts = xts(olliv.df, order.by=olliv.df$Date)
gregor.ts = xts(gregor.df, order.by=gregor.df$Date)

# Merge the 2 XTS Objects by Date
sales.ts = merge(olliv.ts[, 2], gregor.ts[, 2])

# Create dynamic DyGraph for combined sales timeseries
#   Plots of wand sales are labelled by wand maker and color-coded
#   Range selection and mouse highlighting are enabled
#   Years when Voldemort was in power are highlighted
dygraph(sales.ts,
        main='Annual Wand Sales by Ollivander and Gregorovitch for 1970-2017',
        xlab='Year',
        ylab='Wands Sold'
        ) %>%
  dyRangeSelector(height=60) %>%
  dyHighlight(highlightCircleSize=6) %>%
  dySeries(name='Sold_Ollivander', label='Ollivander', color='rgb(198, 97, 25)') %>%
  dySeries(name='Sold_Gregorovitch', label='Gregorovitch', color='rgb(63, 16, 150)') %>%
  dyShading(from="1995-1-1", to='1999-1-1', color='rgb(232, 164, 187)')
  
```