---
title: "Unit 4 Homework"
author: "Dana Geislinger"
date: "January 24, 2018"
output:
  html_document:
    keep_md: true
---
## 1: *FiveThirtyEight* Data Gathering
#### A. Install and load *fivethirtyeight* package
```{r 1a, echo=TRUE}
# Install package
# Previously run: install.packages('fivethirtyeight')
# Load package for use in current session
library(fivethirtyeight)
```

#### B. Assign 18th data set to variable *df*
```{r 1b, echo=TRUE}
# Find 18th data set
data(package="fivethirtyeight")$results[18,]

# Assign 18th data set name to variable 'df'
df <- fivethirtyeight::college_recent_grads
```

#### C. Write the URL of the story related to *df* in a comment
```{r 1c, echo=TRUE}
# Print more detailed list of the 'fivethirtyeight' data sets
vignette("fivethirtyeight", package="fivethirtyeight")

# URL found on vignette webpage
# http://fivethirtyeight.com/features/the-economic-guide-to-picking-a-college-major/
```

#### D. Print dimensions and column names of data.frame *df*
```{r 1d, echo=TRUE}
# Print 'df' dimensions
cat(paste("Rows:",
          dim(df)[1],
          "\nColumns:",
          dim(df)[2],
          sep="\t")
    )
# Print 'df' column names
names(df)
```

## 2: *FiveThirtyEight* Data Analysis
#### A. Write R commands to get *df* column names and number
```{r 2a, echo=TRUE}
# Function to get column names of a data.frame
df.col.names <- function(df) {names(df)}
df.col.names(df)
# Function to get number of columns in a data.frame
df.col.number <- function(df) {dim(df)[2]}
df.col.number(df)
```

#### B. Count occurences of each *major_category* in *df*
```{r 2b, echo=TRUE}
# Generate vector of all unique values in 'major_category' column
unique.categories <- unique(df$major_category)

# Create an empty list to store counts for each major category
category.counts <- list()

# Loop through all major categories
for (category in unique.categories) {
  # Create vector filled with all occurences of 'category' in 'major_category' column of 'df'
  category.occurences <- df$major_category[df$major_category == category]
  # Store the length of that vector in the 'category.counts' list under the correct category name
  category.counts[[category]] <- length(category.occurences)
}

# Convert 'category.counts' to data.frame object and store in new variable 'major_count'
major_count <- data.frame(category.counts, row.names = "Count")

```

#### C. Create barplot for *major_count*
```{r 2c, echo=TRUE}
# Make text appear perpendicular to axis in barplot
par(las=2)

# Create barplot of 'major_count'
#   as.matrix(major_count): convert data.frame object to matrix
#     Allows barplot to use data for 'height'
#   main: title of barplot
#   ylim: vector containing minimum/maximum y-axis values
barplot(as.matrix(major_count),
        main="Number of Undergraduate Majors in each Major Category",
        ylim=c(0, 30)
        )
```

#### D. Write *df* to csv file without row labels
```{r 2d, echo=TRUE}
# Save 'df' data.frame to csv file named 'five_thirty_eight.csv' in current working directory
#   file: name of file to save data to
#   row.names: whether ot not to tell R to save csv file with row labels
write.csv(df,
          file="five_thirty_eight.csv",
          row.names=FALSE
          )
```

## 3. Create MSDS Homework GitHub Repository
#### D. Link to GitHub Repository: https://github.com/danageis/SMU-MSDS-Homework.git