---
title: "Unit 9 Homework"
author: "Dana Geislinger"
date: "March 11, 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### **1.** Harry Potter Cast
IMDb provides a list of all credited cast members in the film **Harry Potter and the Deathly Hallows: Part 2**: http://www.imdb.com/title/tt1201607/fullcredits?ref_=tt_ql_1. This list will be parsed and cleaned using R to present the dataset in a table of actor first/last names and corresponding characer names.

**1a/b:** The table will first be scraped from the webpage and saved in an R *data.frame* object.
```{r 1ab}
# The library 'rvest' will be utilized to parse data from the webpage
require('rvest')

# Extract cast_list table from html parsed from webpage
#   Parse table saved as data.frame 'hp.df'
#   'css_selector' found in Google Chrome's 'Inspect' pane on webpage
hp.url = 'http://www.imdb.com/title/tt1201607/fullcredits?ref_=tt_ql_1'
hp.css_selector = '#fullcredits_content > table.cast_list'
hp.html = read_html(hp.url)
hp.raw_table = html_nodes(hp.html, hp.css_selector)
hp.df = data.frame(html_table(hp.raw_table))

# Verify structure of data.frame object
str(hp.df)
```

**1c:** The raw data parsed from the website is not tidy. The imported data frame *hp.df* will be cleaned in R.
```{r 1c}
# Remove unnecessary columns and rows and define meaningful column names
#   Remove 2 extraneous columns (blank and only containing '...')
#   Remove rows with blank actor names or named 'Rest of cast listed alphebetically:'
hp.df = subset(hp.df, X2 != '' & X2 != 'Rest of cast listed alphabetically:', select=c(X2, X4))
names(hp.df) = c('Actor_Name', 'Character_Name')

# Remove space and newline characters between Character names that span multiple table rows
hp.df$Character_Name = gsub('\\n', '', hp.df$Character_Name)
hp.df$Character_Name = sub('[[:space:]]{2,}', ' ', hp.df$Character_Name)

# Make sure Warwick Davis is credited correctly
subset(hp.df, Actor_Name == "Warwick Davis")
```

**1d/e:** The table on IMDb contains full actor and character names. First/middle and last actor names will be split into separate columns.
```{r 1de}
# The 'extract' function from tidyr will be used to split split the 'Actor_Name' column
require('tidyr')
# Use regular expressions to split actor names and redefine columns in 'hp.df'
hp.df = extract(data=hp.df,
                col=Actor_Name,
                into=c("First_Name", "Last_Name"),
                regex='(^.+) (.+?$)'
                )

# Print Bob Yves Van Hellenberg Hubar's row to make sure middle names were split correctly
subset(hp.df, First_Name == 'Bob Yves Van Hellenberg')

# Print the first 10 rows of the final dataset
head(hp.df, 10)
```

### **2.** SportsBall
Current season statistics for the San Antonio Spurs is available from ESPN: http://www.espn.com/nba/team/stats/_/name/sa/san-antonio-spurs. This data will be parsed to acquire data on current season shooting statistics. Data will be downloaded, cleaned, and presented in R.

**2a/b:** The table will first be scraped from the webpage and saved in an R *data.frame* object.
```{r 2ab}
# Extract Shooting Statistics table from html parsed from webpage
#   Parse table saved as data.frame 'spurs.df'
#   'css_selector' found in Google Chrome's 'Inspect' pane on webpage
spurs.url = 'http://www.espn.com/nba/team/stats/_/name/sa/san-antonio-spurs'
spurs.css_selector = '#my-players-table > div.mod-container.mod-table > div:nth-child(3) > table'
spurs.html = read_html(spurs.url)
spurs.raw_table = html_nodes(spurs.html, spurs.css_selector)
spurs.df = data.frame(html_table(spurs.raw_table))

# Verify structure of raw data set
str(spurs.df)
```

**2c:** Data parsed from ESPN is very messy. Data will be cleaned in R by removing extraneous rows/columns, assigning meaningful column names, creating distinct player name/position columns, and converting improperly formatted data.
```{r 2c}
# Correctly assign column titles
#   Correct column titles were parsed from the webpage, but were erroneously stored in row 2 instead of as column titles
names(spurs.df) = spurs.df[2, ]

# Remove first 2 rows (incorrectly placed titles) and last row (Totals, not necessary for this analysis)
spurs.df = spurs.df[3:19, ]

# Split 'PLAYER' column into 'Name' and 'Position' columns
spurs.df = extract(data=spurs.df,
                   col=PLAYER,
                   into=c('Name', 'Position'),
                   regex='(^.+), (.+$)'
                   )

# Print data types for each variable
sapply(spurs.df, class)

# Convert statistical variables to numeric
spurs.df[, 3:16] = sapply(spurs.df[, 3:16], as.numeric)

# Print data types for dataset variables to confirm changes
sapply(spurs.df, class)

# Print structure of dataset to verify changes
str(spurs.df)
```

**2d:** The potential relationship between player position and field goal percentage per game is of interest. To help illustrate this relationship, a bar graph will be printed using R that prints player field goal percentage color coded by their position.
```{r 2d}
# The ggplot2 library will be used to plot data
require('ggplot2')

# Create ggplot2 horizontal bar chart object of FG% per player color-coded by position
spurs.plot = ggplot(data=spurs.df,
                    aes(x=Name,
                        y=`FG%`,
                        fill=Position)
                    ) + geom_col() + coord_flip()

# Plot barplot with descriptive x and y labels and graph title
spurs.plot + labs(x='Player Name',
                  y='Field Goals Percentage per Game',
                  title='San Antonio Spurs Field Goal Percentage for the 2017-2018 Season'
                  )
```